---
title: "Enrichment of reads mapping based EV and Musa specific genes"
author: "Sadik M"
date: "2023-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r load_packges, echo=FALSE}
library(topGO)
library(GO.db)
library(biomaRt)
BiocManager::install("Rgraphviz")
library(Rgraphviz)
library(tidyverse)
library(ggtext)
library(clusterProfiler)

```

# Load final tidy data containg GO terms, OGs and gene list for all speceis including Musa itinerans 
```{r data}

## data 

GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined <-
  # GO_terms_EV_AED25_MAB_monocots_OGs_genes_blastp %>% 
  PanMusa_AABB_EV_mazia_bedadeti_gene_combined_GO %>%
    filter(str_detect(GO_terms,"GO")) %>%
  mutate(gene_id = case_when(!str_detect(seq.name,"EV") ~ seq.name,
                             TRUE ~ gene_id))

  dir.create(paste0(path_out,"go_terms_enrichment"))
```

# For EV specific genes 

## clusterProfile 
```{r EV_specific_clusProf}

# create placeholder for EV_bedadeti specific ernrichemt outout
EV_specific_clustPro_GOenrich <- c()


  # print(CUTOFF)
  # Extract targe genes set
  
  target_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    filter(str_detect(seq.name, "EV")) %>%
    dplyr::select(gene_id) %>%
    distinct()
  
  
  # Extract global genes set
  GO_temrs_global_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    dplyr::select(GO_terms, gene_id)
  
  
  target_genes_list <-  unique(target_genes$gene_id)
  
  enriched_out <-
    enricher(gene = target_genes_list, TERM2GENE = GO_temrs_global_genes)
  
  if (length(enriched_out) != 0) {
    # convert it to dataframe and tidy-up result
    enriched_out_df <- enriched_out@result %>%
      #separate ratios into 2 columns of data
      separate(BgRatio,
               into = c("size.term", "size.category"),
               sep = "/") %>%
      separate(
        GeneRatio,
        into = c("size.overlap.term", "size.overlap.category"),
        sep = "/"
      ) %>%
      #convert to numeric
      mutate_at(
        vars(
          "size.term",
          "size.category",
          "size.overlap.term",
          "size.overlap.category"
        ),
        as.numeric
      ) %>%
      #Calculate k/K
      mutate("k.K" = size.overlap.term / size.term) %>%
      mutate(frac_cov = CUTOFF,
             category = "EV specific")
    
    
    ## add data
    
    EV_specific_clustPro_GOenrich <-
      enriched_out_df  %>% # add it to your list
      rbind(EV_specific_clustPro_GOenrich)
    
    
  }
  
# remove temporary files

rm(GO_temrs_global_genes,
   enriched_out,
   target_genes_list,
   enriched_out_df)

# tidyup output file

row.names(EV_specific_clustPro_GOenrich) = NULL
```

### ClusterProfiler enrichment plot 
```{r plot_clustProf}
EV_specific_clustPro_GOenrich %>% 
  dplyr::select(-geneID, -Description) %>% 
  dplyr::rename(GO_terms = ID) %>%
  left_join(
    go_odb_aspects_description
  ) %>% 
  mutate(aspects = str_replace_all(aspects, c("biological_process" = "Biological process",
                                              "molecular_function" = "Molecular function",
                                              "cellular_component" = "Cellular component")),
         "log10_pvalue"=log10(p.adjust)) %>% 
  
  # filter(p.adjust <= 0.4) %>% 
  dplyr::filter(description != "NA") %>% 
  dplyr::filter(aspects == 'Biological process') %>%
  dplyr::filter(p.adjust < 0.05) %>%
  ggplot(aes(x=reorder(description, p.adjust), #Reorder gene sets by k/K values
             y=k.K, fill = log10_pvalue)) +
  geom_col( width = 0.8) +
  geom_text(aes(label = Count), hjust=1.1, vjust= 0.5, size=3, fontface = "bold")+
  theme_classic() +
  #Some more customization to pretty it up
  #Flip x and y so long labels can be read
  coord_flip() +
  #fix labels
  labs(y="Proportion of significant genes over<br>the total genes" ,
       x="GO terms description for gene set (Specific to MABS specific)",
  ) +
  scale_fill_gradientn(colours = terrain.colors(10),
                       name  = "log10 pvalue")+
  theme( 
    axis.title.x = element_markdown(size = 9, face = "bold"),
    axis.title.y = element_blank(),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    plot.title = element_markdown(face = "bold"),
    axis.text.y = element_markdown(size = 9,face = "bold"),
    axis.text.x = element_markdown(size = 9),
    legend.title = element_text(size=9),
    legend.text = element_text(size = 8),
    legend.position = "top"
  )
```

# For Musa acuminata specific genes  

## Cluterprofiler enrichment 
```{r MA_specific_clusProf}

# create placeholder for EV_bedadeti specific ernrichemt outout
MA_specific_clustPro_GOenrich <- c()

  # print(CUTOFF)
  # Extract targe genes set
  
  target_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    filter(str_detect(gene_id,"Mac")) %>%
    dplyr::select(gene_id) %>%
    distinct()
  
  
  # Extract global genes set
  GO_temrs_global_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    dplyr::select(GO_terms, gene_id)
  
  
  target_genes_list <-  unique(target_genes$gene_id)
  
  enriched_out <-
    enricher(gene = target_genes_list, TERM2GENE = GO_temrs_global_genes)
  
  if (length(enriched_out) != 0) {
    # convert it to dataframe and tidy-up result
    enriched_out_df <- enriched_out@result %>%
      #separate ratios into 2 columns of data
      separate(BgRatio,
               into = c("size.term", "size.category"),
               sep = "/") %>%
      separate(
        GeneRatio,
        into = c("size.overlap.term", "size.overlap.category"),
        sep = "/"
      ) %>%
      #convert to numeric
      mutate_at(
        vars(
          "size.term",
          "size.category",
          "size.overlap.term",
          "size.overlap.category"
        ),
        as.numeric
      ) %>%
      #Calculate k/K
      mutate("k.K" = size.overlap.term / size.term) %>%
      mutate(
             category = "MA specific")
    
    
    ## add data
    
    MA_specific_clustPro_GOenrich <-
      enriched_out_df  %>% # add it to your list
      rbind(MA_specific_clustPro_GOenrich)
    
    
  }
  
  

# remove temporary files

rm(GO_temrs_global_genes,
   enriched_out,
   target_genes_list,
   enriched_out_df)

# tidyup output file

row.names(MA_specific_clustPro_GOenrich) = NULL

```

### Plot clusterProfiler enrichment
```{r plot_clustProf}
MA_specific_clustPro_GOenrich %>% 
  dplyr::select(-geneID, -Description) %>% 
  dplyr::rename(GO_terms = ID) %>%
  left_join(
    go_odb_aspects_description
  ) %>% 
  mutate(aspects = str_replace_all(aspects, c("biological_process" = "Biological process",
                                              "molecular_function" = "Molecular function",
                                              "cellular_component" = "Cellular component")),
         "log10_pvalue"=log10(p.adjust)) %>% 
  
  # filter(p.adjust <= 0.4) %>% 
  dplyr::filter(description != "NA") %>% 
  dplyr::filter(aspects == 'Biological process') %>%
  dplyr::filter(p.adjust < 0.05) %>%
  ggplot(aes(x=reorder(description, p.adjust), #Reorder gene sets by k/K values
             y=k.K, fill = log10_pvalue)) +
  geom_col( width = 0.8) +
  geom_text(aes(label = Count), hjust=1.1, vjust= 0.5, size=3, fontface = "bold")+
  theme_classic() +
  coord_flip() +
  #fix labels
  labs(y="Proportion of significant genes over<br>the total genes" ,
       x="GO terms description for gene set (Specific to MABS specific)",
       #title = "Musa species specific" 
  ) +
  scale_fill_gradientn(colours = terrain.colors(10),
                       name  = "log10 pvalue")+
  
  # scale_fill_continuous(name  = "log10 pvalue")+
  theme( 
    axis.title.x = element_markdown(size = 9, face = "bold"),
    axis.title.y = element_blank(),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    plot.title = element_markdown(face = "bold"),
    axis.text.y = element_markdown(size = 9,face = "bold"),
    axis.text.x = element_markdown(size = 9),
    legend.title = element_text(size=9),
    legend.text = element_text(size = 8),
    legend.position = "top"
  )
```

# For Musa balbisiana specific genes  

## Cluterprofiler enrichment 
```{r MA_specific_clusProf}

# create placeholder for EV_bedadeti specific ernrichemt outout
MB_specific_clustPro_GOenrich <- c()

  # print(CUTOFF)
  # Extract targe genes set
  
  target_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    filter(str_detect(gene_id,"Mba")) %>%
    dplyr::select(gene_id) %>%
    distinct()
  
  
  # Extract global genes set
  GO_temrs_global_genes <-
    GO_temrs_PanReadsMapping_EV_mazia_bedadeti_gene_combined %>%
    dplyr::select(GO_terms, gene_id)
  
  
  target_genes_list <-  unique(target_genes$gene_id)
  
  enriched_out <-
    enricher(gene = target_genes_list, TERM2GENE = GO_temrs_global_genes)
  
  if (length(enriched_out) != 0) {
    # convert it to dataframe and tidy-up result
    enriched_out_df <- enriched_out@result %>%
      #separate ratios into 2 columns of data
      separate(BgRatio,
               into = c("size.term", "size.category"),
               sep = "/") %>%
      separate(
        GeneRatio,
        into = c("size.overlap.term", "size.overlap.category"),
        sep = "/"
      ) %>%
      #convert to numeric
      mutate_at(
        vars(
          "size.term",
          "size.category",
          "size.overlap.term",
          "size.overlap.category"
        ),
        as.numeric
      ) %>%
      #Calculate k/K
      mutate("k.K" = size.overlap.term / size.term) %>%
      mutate(
             category = "MA specific")
    
    
    ## add data
    
    MB_specific_clustPro_GOenrich <-
      enriched_out_df  %>% # add it to your list
      rbind(MB_specific_clustPro_GOenrich)
    
    
  }
  
  

# remove temporary files

rm(GO_temrs_global_genes,
   enriched_out,
   target_genes_list,
   enriched_out_df)

# tidyup output file

row.names(MB_specific_clustPro_GOenrich) = NULL

```

### Plot clusterProfiler enrichment
```{r plot_clustProf}
MB_specific_clustPro_GOenrich %>% 
  dplyr::select(-geneID, -Description) %>% 
  dplyr::rename(GO_terms = ID) %>%
  left_join(
    go_odb_aspects_description
  ) %>% 
  mutate(aspects = str_replace_all(aspects, c("biological_process" = "Biological process",
                                              "molecular_function" = "Molecular function",
                                              "cellular_component" = "Cellular component")),
         "log10_pvalue"=log10(p.adjust)) %>% 
  
  # filter(p.adjust <= 0.4) %>% 
  dplyr::filter(description != "NA") %>% 
  dplyr::filter(aspects == 'Biological process') %>%
  dplyr::filter(p.adjust < 0.05) %>%
  ggplot(aes(x=reorder(description, p.adjust), #Reorder gene sets by k/K values
             y=k.K, fill = log10_pvalue)) +
  geom_col( width = 0.8) +
  geom_text(aes(label = Count), hjust=1.1, vjust= 0.5, size=3, fontface = "bold")+
  theme_classic() +
  coord_flip() +
  #fix labels
  labs(y="Proportion of significant genes over<br>the total genes" ,
       x="GO terms description for gene set (Specific to MABS specific)",
       #title = "Musa species specific" 
  ) +
  scale_fill_gradientn(colours = terrain.colors(10),
                       name  = "log10 pvalue")+
  
  # scale_fill_continuous(name  = "log10 pvalue")+
  theme( 
    axis.title.x = element_markdown(size = 9, face = "bold"),
    axis.title.y = element_blank(),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    plot.title = element_markdown(face = "bold"),
    axis.text.y = element_markdown(size = 9,face = "bold"),
    axis.text.x = element_markdown(size = 9),
    legend.title = element_text(size=9),
    legend.text = element_text(size = 8),
    legend.position = "top"
  )
```

# save output to disk 
```{r}
# ls(pattern = "_topGO")

EV_specific_clustPro_GOenrich %>%
  rbind(
    MAB_specific_clustPro_GOenrich
  ) %>%
  write.table(
    paste0(
      path_out,
      "/go_terms_enrichment/",
      "EV_MAB_clusterPr.readsMapping_enrichment.txt"
    ),
    col.names = T,
    row.names = F,
    quote = F,
    sep = '\t'
  )

```

# save project
```{r save_project}

save.image(file = "cluterProfile_enrichment.RData")

# load(file = "../monocot_blastp_all.RData")

```





