# Load packages

```{r load_packges }
library(topGO)
# library(GO.db)
library(biomaRt)
library(Rgraphviz)
library(tidyverse)
library(ggtext)
```

# Load final tidy data containg GO terms, OGs and gene list for all speceis including Musa itinerans 
```{r }

library(tidyverse)

GO_terms_geneID_EV_AED25 <-
  GO_terms_EV_AED25_TE %>% 
  filter(GO_terms != "NA") %>%
  filter(!str_detect(seq.name,"^M")) %>%
  select(GO_terms, seq.name)

GO_terms_geneID_EV_AED25 %>% filter(!str_detect(GO_terms,"GO"))


go_odb_aspects_description <- read.delim("go_terms/EV_MAB_enrichment/go_odb_aspects_description.txt") 


# GO_terms_EV_AED25_OGs_genes from combine_orthogroup_GO_terms
```
  

# Orthogroups present in either enset landrace but missing in Musa

```{r enset_specific}

dir.create("orthofinder/go_terms_enrichment")

EV_specific_MAB <- 

  # EV_bedadeti_specific_MAB <-
  GO_terms_EV_AED25_MAB_OGs_genes  %>% 
  left_join(
    EV_AED25_MAB.OG_GeneCount %>% 
      filter(
        Ensete_ventricosum_bedadeti == 1,
        Ensete_ventricosum_mazia == 0,
        
        Musa_acuminata == 0,
        Musa_balbisiana == 0
      ) %>% select(Orthogroup) %>% 
      mutate(EV_shared_GF = 'yes') 
  ) %>% 
  mutate(
    EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
  ) %>%
    filter(EV_shared_GF != "NA") %>%
  filter(GO_terms != "NA") %>%
  
  select(seq.name) %>%
  rbind( 

    # mazia specific
  GO_terms_EV_AED25_MAB_OGs_genes  %>% 
  left_join(
    EV_AED25_MAB.OG_GeneCount %>% 
      filter(
        Ensete_ventricosum_bedadeti == 0,
        Ensete_ventricosum_mazia == 1,
        
        Musa_acuminata == 0,
        Musa_balbisiana == 0
      ) %>% select(Orthogroup) %>% 
      mutate(EV_shared_GF = 'yes') 
  ) %>% 
  mutate(
    EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
  ) %>%
  filter(EV_shared_GF != "NA") %>%
  filter(GO_terms != "NA") %>%
  
  select(seq.name),

  # shared EV 

GO_terms_EV_AED25_MAB_OGs_genes  %>%
filter(GO_terms != "NA") %>%
left_join(
    EV_AED25_MAB.OG_GeneCount %>%
      filter(
        Ensete_ventricosum_bedadeti == 1,
        Ensete_ventricosum_mazia == 1,
        
        Musa_acuminata == 0,
        Musa_balbisiana == 0
        
      ) %>% 
      select(Orthogroup) %>%
      mutate(EV_shared_GF = 'yes')
  ) %>%
  # mutate(
  #   EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
  # ) %>%
  filter(EV_shared_GF == "yes") %>%
  filter(GO_terms != "NA")  %>%
  select(seq.name)) %>%
  distinct()


# TopGo Enrichment 

## topGO
```{r enrichment }
## Create a list of containing genes of interest 

targe_gene_list=EV_specific_MAB$seq.name

## Create a list containing all genes that we want to compare out genes of interest.

all_genes_MAB_list = GO_terms_geneID_EV_AED25_MAB$seq.name

#Step 2: Create GO annotation

all_genes_EV_AED25_EG_MABS_db <- 
  GO_terms_geneID_EV_AED25_MAB %>% 
  left_join(go_odb_aspects_description) %>%
  filter(aspects == "biological_process") %>%
  dplyr::select(-description,-defination) 

# build the gene 2 GO annotation list (needed to create topGO object)
gene_2_GO=unstack(all_genes_EV_AED25_EG_MABS_db[,c(1,2)])

# remove any candidate genes without GO annotation
keep = targe_gene_list %in% all_genes_EV_AED25_EG_MABS_db[,2]$seq.name
keep =which(keep==TRUE)
targe_gene_list=targe_gene_list[keep]

# make named factor showing which genes are of interest
geneList=factor(as.integer(all_genes_MAB_list %in% targe_gene_list))
names(geneList)= all_genes_MAB_list

#Step 3: Make topGO data object
GOdata=new('topGOdata', ontology='BP', allGenes = geneList, annot = annFUN.gene2GO, gene2GO = gene_2_GO)

# Step 4: Test for significance
# define test using the classic algorithm with fisher (refer to [1] if you want to understand how the different algorithms work)
classic_fisher_result=runTest(GOdata, algorithm='classic', statistic='fisher')


# define test using the weight01 algorithm (default) with fisher
weight_fisher_result=runTest(GOdata, algorithm='weight01', statistic='fisher') 

# generate a table of results: we can use the GenTable function to generate a summary table with the results from tests applied to the topGOdata object.
allGO=usedGO(GOdata)
all_res=GenTable(GOdata, weightFisher=weight_fisher_result, orderBy='weightFisher', topNodes=length(allGO))

# Step 5: Correcting for multiple testing

#performing BH correction on our p values
p.adj=round(p.adjust(all_res$weightFisher,method="BH"),digits = 4)

# create the file with all the statistics from GO analysis
all_res_final=cbind(all_res,p.adj)
all_res_final=all_res_final[order(all_res_final$p.adj),]

all_res_final <- all_res_final %>%
  mutate(p.adj = str_replace_na(p.adj,"0.0000001"),
         p.adj = as.numeric(p.adj)) 


# set all near zeros to 0.0000001
all_res_final$p.adj[all_res_final$p.adj==0.00e+00] <- 0.0000001

# p.adj <= 0.05 
p.adj_cuttoff <- 0.05

EV_specific_MAB_topGO <-  
  all_res_final[all_res_final$p.adj<=p.adj_cuttoff,c("GO.ID","Term","Annotated","Significant", "Expected", "p.adj")]


EV_specific_MAB_topGO$Category <- "Present in either EV & absent in Musa"

# save 
# write.table(EV_bedadet_MAB,file = paste0("../../reannotation_analysis/mazia_reannotation/orthofinder/go_terms_enrichment/EV_AED25_EG_MABS/EV_bedadet_MAB.txt"),
#             col.names = F, row.names = T, quote = F, sep = '\t')



# PLOT the GO hierarchy plot: the enriched GO terms are colored in yellow/red according to significance level

# pdf(file='topGOPlot_fullnames.pdf', height=12, width=12, paper='special', pointsize=18)
# showSigOfNodes(GOdata, score(weight_fisher_result), useInfo = "none", sigForAll=FALSE, firstSigNodes=5,.NO.CHAR=50)
# printGraph(GOdata, classic_fisher_result, firstSigNodes =6, fn.prefix = "../../reannotation_analysis/mazia_reannotation/orthofinder/go_terms_enrichment/AED25/EV_shared_MAB_topGO", useInfo = "def")

# remove temp files
rm(classic_fisher_result,GOdata,gene_2_GO,weight_fisher_result,all_res_final,p.adj,
   all_res,all_res_final.padj_05,all_res_final,geneList,targe_gene_list,all_genes_MAB_list,keep,allGO)

# dev.off()

```


# Orthogroups present in at least one Musa and absent from EV

```{r Musa_specic}    
# Specific to MUSA genomes

MUSA_MAB_specific <- 
  GO_terms_EV_AED25_MAB_OGs_genes  %>%
  left_join(
    EV_AED25_MAB.OG_GeneCount %>%
      filter(
        Ensete_ventricosum_bedadeti == 0,
        Ensete_ventricosum_mazia == 0,
        
        Musa_acuminata == 1,
        Musa_balbisiana == 1,

      ) %>% select(Orthogroup) %>%
      mutate(EV_shared_GF = 'yes')
  ) %>%
  mutate(
    EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
  ) %>%
  filter(EV_shared_GF != "NA") %>%
  filter(GO_terms != "NA") %>%
  select(seq.name) %>%
  distinct() %>%
  rbind(
    # Specific to  MB, MS, MI
    # Specific to MA, MS, MI
    
    GO_terms_EV_AED25_MAB_OGs_genes  %>%
      left_join(
        EV_AED25_MAB.OG_GeneCount %>%
          filter(
            Ensete_ventricosum_bedadeti == 0,
            Ensete_ventricosum_mazia == 0,
            
            Musa_acuminata == 1,
            Musa_balbisiana == 0,
                        
          ) %>% select(Orthogroup) %>%
          mutate(EV_shared_GF = 'yes')
      ) %>%
      mutate(
        EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
      ) %>%
      filter(EV_shared_GF != "NA") %>%
      filter(GO_terms != "NA") %>%
      select(seq.name) %>%
      distinct(),
    
    # Specific to MA, MB, MI 
    GO_terms_EV_AED25_MAB_OGs_genes  %>%
      left_join(
        EV_AED25_MAB.OG_GeneCount %>%
          filter(
            Ensete_ventricosum_bedadeti == 0,
            Ensete_ventricosum_mazia == 0,
            
            Musa_acuminata == 0,
            Musa_balbisiana == 1,
            
          ) %>% select(Orthogroup) %>%
          mutate(EV_shared_GF = 'yes')
      ) %>%
      mutate(
        EV_shared_GF = str_replace_na(EV_shared_GF,"NA")
      ) %>%
      filter(EV_shared_GF != "NA") %>%
      filter(GO_terms != "NA") %>%
      select(seq.name) %>%
      distinct()
  ) %>% 
  distinct() 
```

# TopGO enrichment using same script changing the input file


# save project
```{r save_project}
save.image(file = "orthofinder/EV_AED25_MAB/OrthoFinder.EV.AED25.MAB.enrich.RData")
```


